---
title: R Scrollytelling Template Project
author: IMERSS biodiversity informatics working group
date: December 30th, 2022
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---
## Introduction

This project hosts a skeleton structure suitable for building out and publishing a map-based scrollytelling
interface based on R Markdown. You can build your own scrollytelling interface using R and R Markdown,
using free tools such as git, R and R studio, and publish and host it for free using [GitHub Pages](https://pages.github.com/).

This first section renders a basic vector map layer represented as an ESRI Shapefile, perhaps
[exported from a GIS system such as QGIS](https://docs.qgis.org/3.22/en/docs/user_manual/managing_data_source/supported_data.html#esri-shapefile-format).

```{r, echo=FALSE, message=FALSE}
source("scripts/vector_map_script.R")
baseMap
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
