---
title: Átl’ḵa7tsem/Howe Sound Biosphere Vegetation Analysis
author: Antranig Basman, Judith Holm, Murray Journeay & Andrew Simon
date: March 26th, 2023
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
plants <- read.csv("tabular_data/Howe_Sound_vascular_plant_records_consolidated.csv")
plant.summary <- read.csv("tabular_data/vascular_plant_summary_resynthesized_2023-03-05.csv")
new <- plant.summary %>% filter(str_detect(reportingStatus, "new")) 
confirmed <- plant.summary %>% filter(reportingStatus == 'confirmed') 
reported <- plant.summary %>% filter(reportingStatus == 'reported') 
observed <- plant.summary %>% filter(iNatObservationStatus == 'observed') 
```

## Introduction

## Howe Sound's vascular plant diversity

Over one thousand vascular plant species are known to occur within the Átl’ḵa7tsem/Howe Sound Biosphere, represented by `r nrow(plants)` occurrence records. This diverse flora spans nine biogeoclimatic units, from lowland coastal western hemlock forests and associated woodland communities to high elevation subalpine and alpine plant communities.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
source("scripts/Vascular_plant_map_script.R")
speciesMap
```
*Map 1. Ecological communities*

Explore the region's flora by selecting from the biogeoclimatic units displayed on the map:

* Alpine Tundra zone ([CMAunp](#region:CMAunp))
* Mountain Hemlock zone ([MHmm1](#region:MHmm1), [MHmm2](#region:MHmm2))
* Coastal Western Hemlock zone ([CWHdm](#region:CWHdm), [CWHds1](#region:CWHds1), [CWHms1](#region:CWHms1), [CWHvm1](#region:CWHvm1), [CWHvm2](#region:CWHvm2), [CWHxm1](#region:CWHxm1))

## History of collection activities within the region

While Indigenous knowledge of Átl’ḵa7tsem's plant life traces back to time immemorial, Western scientific documentation of the region's flora dates to 1897, with the early collection efforts of James Anderson and Albert J. Hill. Over the last century and a half, efforts to document the regional flora were most concentrated in the 1920s, and from the 1960s through the 1980s, before expanding significantly in the 21st century with the advent of iNaturalist, with many thousands of observations made by community members through the <a href = "https://www.squamishenvironment.ca/projects/biodiversity-squamish/">Howe Sound Biodiversity project</a>.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# According to https://stackoverflow.com/questions/5031630/how-to-source-r-file-saved-using-utf-8-encoding this strange strategy is now necessary since the 
# file contains a character outside the Unicode BMP
eval(parse("scripts/History_of_collection_activities.R", encoding="UTF-8"))
heatMap
speciesPlot
# animate(
#    plot = plot,
#    render = gifski_renderer(),
#    height = 420,
#    width = 700, 
#    duration = 10,
#    fps = 30,
#    res = 100,
#    end_pause = 120)
```
*Map 2. Historical collection activities*
*Figure 1. Historical collection activities*

## Community science contributions

Through the combined efforts of `r length(unique(plants$recordedBy))` individuals, the Átl’ḵa7tsem/Howe Sound naturalist community has documented `r nrow(plant.summary)` species, confirming `r round((nrow(confirmed)/(nrow(confirmed)+nrow(reported))*100), digits = 0)` percent of the historically reported vascular plant diversity and adding `r nrow(new)` new species reports to the list. Community members have now documented `r round((nrow(observed)/(nrow(plant.summary))*100), digits = 0)` percent of the plant diversity known to the region on iNaturalist.

```{r, echo=FALSE, message=FALSE}
source("scripts/Community_science_contributions.R")
reportingStatusMap
reportingStatusFig
```
*Map 3. Species Reporting Status*
*Figure 2: Species Reporting Status*

## Top contributions to our knowledge of the regional plant diversity

Integrate some of Ben Keen's code to show photos some of the most commonly documented plants, recent observations, top observers?

## Vascular plant diversity within Howe Sound's protected areas

```{r, echo=FALSE, message=FALSE, warning=FALSE}
source("scripts/Protected_areas.R")

protectedAreaMap
protected.area.plot 
```
*Map 4. Protected Areas*
*Figure 3. Species Reported by Protected Area*

Summary of plant diversity documented within protected areas; call to action

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
