---
title: Átl’ka7tsem/Howe Sound Biosphere Vegetation Analysis
author: Andrew Simon, Antranig Basman & Murray Journeay
date: January 9th, 2022
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 plants <- read.csv("tabular_data/Howe_Sound_vascular_plant_records_consolidated.csv")
```

## Introduction

## Howe Sound's vascular plant diversity

Over one thousand vascular plant species are known to occur within the Átl’ka7tsem/Howe Sound Biosphere, represented by `r nrow(plants)` species occurrence records. This diverse flora spans 11 biogeoclimatic units, from lowland coastal western hemlock forests and associated woodland communities to high elevation subalpine and alpine plant communities.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
source("scripts/Vascular_plant_map_script.R")
speciesMap
```
### Map 1. Ecological communities

Explore the region's flora by selecting from the biogeoclimatic units displayed on the map:

* Alpine (CMA)
* Engelmann spruce subalpine fir communities (ESSFmw)
* Mountain hemlock communities (MHmm)
* Coastal western hemlock communities (CWHxm,vm,ms,ds,dm)

[Data visualization similar to Xetthecum but summarizing plant diversity by BEC unit with a simplified interface showing just the map, the (selectable) communities, and associated species lists below, including a small image of the species]

## History of collection activities within the region

While Indigenous knowledge of Átl’ka7tsem's plant life traces back to time immemorial, Western scientific documentation of the region's flora dates to 1897, with the early collection efforts of James Anderson and Albert J. Hill. Over the last century and a half, the activities of botanists increased in the 1920s, and from the 1960s through the 1980s, before massively expanding in the 21st century with the advent of iNaturalist, with many thousands of observations made by community members through the <a href = "https://www.squamishenvironment.ca/projects/biodiversity-squamish/">Howe Sound Biodiversity project</a>.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
source("scripts/History_of_collection_activities.R", encoding="UTF-8")
heatMap
speciesPlot
# animate(
#    plot = plot,
#    render = gifski_renderer(),
#    height = 420,
#    width = 700, 
#    duration = 10,
#    fps = 30,
#    res = 100,
#    end_pause = 120)
```
\

### Figure 1. Historical collection activities

[Map cycles through twenty-five decades of search effort, synchronized with a graph showing new species reported over time (choropleth).]

## Community science contributions

Through the combined efforts of XXXX individuals using iNaturalist, the Átl’ka7tsem/Howe Sound community has documented XXX species, confirming XXX percent of the historically reported vascular plant diversity and adding XX new species reports to the list.

[Summary statistics similar to those displayed at www.biogaliano.org/stats]

```{r, echo=FALSE, message=FALSE}
# source("scripts/Community_science_contributions.R")
# speciesMap
```


### Figure 2: Pie charts / bar plots summarizing how well documented plant families are (e.g., Bio Galiano stats)

Select each plant family to track our progress documenting the regional flora. Unconfirmed historical reports are in green; confirmed historical reports in orange; and new records in blue.

### Can we integrate some of Ben Keen's code to show photos some of the most commonly documented plants, recent observations, top observers?


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
