---
title: Xetthecum Cultural Mapping Project
author: IMERSS biodiversity informatics working group
date: December 20th, 2023
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---
## Introduction

Welcome to Xetthecum (in Hul’q’umi’num’ followed by English), an area on Galiano Island BC that is of ecological and cultural significance. Complex in its ecology, cultural history, and contemporary land-use, the boundary of Xetthecum is roughly delimited by the extent of the [Greig Creek](#region:GreigCreek) watershed, including the watercourse descending from [Laughlin Lake](#region:LaughlinLake) to [Retreat Cove](#region:RetreatCove), spanning residential and agricultural lands, protected and covenanted areas, and a public shore access.

```{r, echo=FALSE, message=FALSE}
source("scripts/vector_map_script.R")
introMap <- mx_intro_map()
introMap
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Caves

![Spiritual cave Shhwuli’wun - praying for the spirit of the deer
](img/CaveIllustration.png)


Florence’s grandpa said the cave: if you take care of the deer when you kill a deer for food, its spirit will come and go; if you respect it when you kill it, it will always come back; respect the body and its parts; make use of its skin; the hooves and the skin were even used in the Big House.

Deer Island (Parker):
There's a mention that Parker Island was known as 'Deer Island' where deer ceremonies occurred.

Spiritual Significance:
Florence mentioned that her grandfather used a specific cave called Shhwuli’wun as a sacred place for praying, particularly for the spirit of the deer.
She recalled feeling emotional at this cave, mentioning that after her experience there, her mother passed away.

Ceremonial Importance:
Retreat Cove, including its caves, served as locations for private ceremonies within the Penelakut community.
Florence emphasized the significance of these caves for specific rituals, including training for ceremonies such as puberty rites.

Protective Measures:
There were efforts made by someone named Ken to protect a hidden cave called Shqula’tsus, suggesting a community concern for safeguarding these culturally important sites.

Vandalism Concerns:
Florence expressed concern about ongoing issues related to vandalism and damage occurring at the caves, highlighting the need to protect and respect these important cultural spaces.

Shqula’tsus —”a refuge, was protected by Ken (it is not on a map), a hidden cave”

```{r, echo=FALSE, message=FALSE}

```
